### 2013-07-24 doc/asplos14/communication.tex
more small fixes

Each processing core in a given system node is responsible for
aggregating and sending the resulting messages from all processors on
that node to a set of destination nodes.  Each core periodically
executes a task responsible for sending messages.  This task examines
the private (to each core) message lists for each destination node it
is responsible for managing and, if the list is long enough or a
message has waited past a time-out period, all messages to a given
destination system node from that source system node are sent.
Aggregating and sending a message involves manipulating a set of
shared data-structures (the message lists). This is done using CAS
(compare-and-swap) operations to avoid high synchronization costs.
Note that we use a per-processor array of message lists
that is only periodically modified across processor cores because we
empirically determined this approach was faster (sometimes
significantly) than a global per-system node array of message lists.

>>>>

Each processing core in a given system node is responsible for
aggregating and sending the resulting messages from all processors on
that node to a set of destination nodes.  Each core periodically
executes a task responsible for sending messages.  This task examines
the private (to each core) message lists for each destination node it
is responsible for managing and, if the list is long enough or a
message has waited past a time-out period, all messages to a given
destination system node from that source system node are sent.
Aggregating and sending a message involves manipulating a set of
shared data-structures (the message lists). This is done using CAS
(compare-and-swap) operations to avoid high synchronization costs.
Note that we use a per-processor array of message lists
that is only periodically modified across processor cores after
experimentally determining that this approach was faster (sometimes
significantly) than a global per-system node array of message lists.

### 2013-07-24 doc/asplos14/evaluation.tex
more small fixes

Figure~\ref{fig:context-switch-exp} shows the average context switch
time as the number of workers grow. At our standard operating point
($\approx$1K workers), context switch time is on the order of
50~ns. As we add workers, the time increases slowly, but levels off:
we also ran with 500,000 workers (10~times what is shown in the
figure) and found that context switch time was around 75~ns. In
comparison, for the same yield test using kernel-level Pthreads on a
single core, the switch time is 450~ns for a few threads and 800~ns for
1000--32000 threads.

If we calculate aggregate context switch \emph{rate\/} of all
cores in a node, we find that with prefetching, \Grappa context
switching is
limited \emph{not\/} by memory latency, as normally assumed, but rather memory bandwidth.
Specifically, we empirically found that 4~cache lines (1~for worker
struct and 3 for stack data) was sufficient to avoid cache misses in
the microbenchmark. Every context switch then requires 8~cache-line transfers. The off-chip
bandwidth of a single socket in our system is 270M cache lines per
second~\cite{porterfield:bw,Nelson:hotpar11}. This implies that, in the limit,
we can sustain at most 34M context switches per second per socket (a context-switch time of 29~ns).

@@

We look at two classes of trees, T1 and T3, from
the original benchmark. T1 trees are very shallow and wide (i.e., significant
parallel slack \TODO{cite Valiant and say more somewhere about BSP}), while T3 trees are very deep (i.e., little parallel slack).
When the access to each vertex is a random access, the critical path to search
T3 trees is very long, hence the low performance and scalability. On such
trees, we do not expect there to be sufficient concurrency for any system,
including \Grappa, to achieve high throughput -- at the 16-node data point,
the average active tasks per core over the search is 775 for T1 and 13 for T3.
On the other hand, \Grappa performs and scales very well for T1
trees.

@@

In UTS, we obtained the best throughput by flushing frequently: on the
order of 100~$\mu$s. This, combined with the fewer number of remote
accesses, means that there are a limited number of requests generated
per flush interval, leading to packets primarily smaller than~10KB.

In BFS, we see a trimodal distribution. There are two reasons for
this. First, in BFS, we obtained the best throughput by flushing at
least 500~$\mu$s apart, and also flushing whenever a core is
idle. Since this process ignores whether another flush has occurred
recently, it leads to a large number of small messages. This leads to
the first and smallest mode of the distribution.  The other two modes
are due to a locality optimization in the \Grappa version of BFS. To
minimize unnecessary remote communication, each core visits vertices
whose edge lists are stored locally. Since we are traversing power-law graphs, a small fraction of vertices have a very high number of edges. This means that occasionally a core visits a high-degree vertex which allows it to very quickly generate a large batch of requests. This interacts well with aggregation; when we find high degree vertices, we are able to aggregate packets on the order of 40KB. In flush intervals without high-degree vertices, we see packets on the order of 8KB.

>>>>

Figure~\ref{fig:context-switch-exp} shows the average context switch
time as the number of workers grow. At our standard operating point
($\approx$1K workers), context switch time is on the order of
50ns. As we add workers, the time increases slowly, but levels off:
we also ran with 500,000 workers (10~times what is shown in the
figure) and found that context switch time was around 75ns. In
comparison, for the same yield test using kernel-level Pthreads on a
single core, the switch time is 450ns for a few threads and 800ns for
1000--32000 threads.

If we calculate aggregate context switch \emph{rate\/} of all
cores in a node, we find that with prefetching, \Grappa context
switching is
limited \emph{not\/} by memory latency, as normally assumed, but rather memory bandwidth.
Specifically, we empirically found that 4~cache lines (1~for worker
struct and 3 for stack data) was sufficient to avoid cache misses in
the microbenchmark. Every context switch then requires 8~cache-line transfers. The off-chip
bandwidth of a single socket in our system is 270M cache lines per
second~\cite{porterfield:bw,Nelson:hotpar11}. This implies that, in the limit,
we can sustain at most 34M context switches per second per socket (a context-switch time of 29ns).

@@

We look at two classes of trees, T1 and T3, from
the original benchmark. T1 trees are very shallow and wide (i.e., significant
parallel slack \TODO{cite Valiant and say more somewhere about BSP}), while T3 trees are very deep (i.e., little parallel slack).
When the access to each vertex is a random access, the critical path to search
T3 trees is very long, hence the low performance and scalability. On such
trees, we do not expect there to be sufficient concurrency for any system,
including \Grappa, to achieve high throughput -- at the 16-node data point,
the average active tasks per core over the search is 775 for T1 and
only 13 for T3.
On the other hand, \Grappa performs and scales very well for T1
trees.

@@

In UTS, we obtained the best throughput by flushing frequently: on the
order of 100$\mu$s. This, combined with the fewer number of remote
accesses, means that there are a limited number of requests generated
per flush interval, leading to packets primarily smaller than~10KB.

In BFS, we see a trimodal distribution. There are two reasons for
this. First, in BFS, we obtained the best throughput by flushing at
least 500$\mu$s apart, and also flushing whenever a core is
idle. Since this process ignores whether another flush has occurred
recently, it leads to a large number of small messages. This leads to
the first and smallest mode of the distribution.  The other two modes
are due to a locality optimization in the \Grappa version of BFS. To
minimize unnecessary remote communication, each core visits vertices
whose edge lists are stored locally. Since we are traversing power-law graphs, a small fraction of vertices have a very high number of edges. This means that occasionally a core visits a high-degree vertex which allows it to very quickly generate a large batch of requests. This interacts well with aggregation; when we find high degree vertices, we are able to aggregate packets on the order of 40KB. In flush intervals without high-degree vertices, we see packets on the order of 8KB.

### 2013-07-24 doc/asplos14/intro.tex
more small fixes

\vspace{0.5ex}
\noindent{\bf Poor data locality and frequent communication.} Data reference patterns of irregular applications are unpredictable and tend to be spread across the entire system. This results in frequent requests for small pieces of remote data. Caches are of little assistance because of low temporal and spatial locality. Prefetching is of limited value because request locations are not known early enough. Data-parallel frameworks such as MapReduce~\cite{Dean:2008:MSD:1327452.1327492} are ineffective because they rely on effective data partitioning and regular communication patterns. Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.

\vspace{0.5ex} \noindent{\bf High network communication latency.} The performance challenges of frequent communication are exacerbated by high network latency relative to processor performance. Latency of commodity networks runs anywhere from a few to hundreds of microseconds -- tens of thousands of processor clock cycles.  Since irregular application tasks encounter remote references dynamically during execution and must resolve them before making further progress, stalls are frequent and lead to severely underutilized compute resources.

>>>>

\vspace{0.5ex}
\noindent{\bf Poor data locality and frequent communication} Data reference patterns of irregular applications are unpredictable and tend to be spread across the entire system. This results in frequent requests for small pieces of remote data. Caches are of little assistance because of low temporal and spatial locality. Prefetching is of limited value because request locations are not known early enough. Data-parallel frameworks such as MapReduce~\cite{Dean:2008:MSD:1327452.1327492} are ineffective because they rely on effective data partitioning and regular communication patterns. Consequently, commodity networks, which are designed for large packets, achieve just a fraction of their peak bandwidth on small messages, starving application performance.

\vspace{0.5ex} \noindent{\bf High network communication latency} The performance challenges of frequent communication are exacerbated by high network latency relative to processor performance. Latency of commodity networks runs anywhere from a few to hundreds of microseconds -- tens of thousands of processor clock cycles.  Since irregular application tasks encounter remote references dynamically during execution and must resolve them before making further progress, stalls are frequent and lead to severely underutilized compute resources.

### 2013-07-24 doc/asplos14/method.tex
more small fixes

We implemented the \Grappa in C++ for the Linux operating system. The core
runtime system system is about 15K lines of code. We ported a number of
benchmarks to \Grappa as well as collected and optimized a set of comparison
benchmarks for XMT, MPI, and UPC~\cite{UPC}. We run the \Grappa, MPI and UPC
experiments on a cluster of AMD Interlagos processors. Nodes have 32 2.1-GHz
cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand
network cards. Nodes are connected via a QLogic InfiniBand switch. We also
compare \Grappa to a 128-node Cray XMT (3rd generation MTA). Each node
consists of a 500-MHz MTA Threadstorm multithreaded processor that supports
128~streams. The machine uses Cray's proprietary SeaStar2 interconnection
network.

@@

\vspace{0.5ex}
\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the Cray XMT's early predecessor once held the world speed record~\cite{TeraRecord}. The largest problem size, class ``D'', ranks two billion uniformly distributed random integers using either a bucket or a counting sort algorithm, depending on the strengths of the system. Bucket sort executes a greater number of loops, but is able to leverage locality and avoid communication completely in the final phase, ranking within buckets. For these reasons, the MPI reference version and our \Grappa implementation use bucket sort. On the other hand, the Cray XMT cannot take advantage of locality, but has an efficient compiler-supported parallel prefix sum, so it performs best using the counting sort algorithm.  The performance metrics for NAS Parallel Benchmarks, including IntSort, are ``millions of operations per second'' (MOPS). For IntSort, this ``operation'' is ranking a single key, so it is roughly comparable to ``GUPS'' or ``TEPS.''

>>>>

We implemented the \Grappa in C++ for the Linux operating system. The core
runtime system system is about 15K lines of code. We ported a number of
benchmarks to \Grappa as well as collected and optimized a set of comparison
benchmarks for XMT, MPI, and UPC~\cite{UPC}. We ran the \Grappa, MPI, and UPC
experiments on a cluster of AMD Interlagos processors. Nodes have 32 2.1-GHz
cores in two sockets, 64GB of memory, and 40Gb Mellanox ConnectX-2 InfiniBand
network cards. Nodes are connected via a QLogic InfiniBand switch. We also
compare \Grappa to a 128-node Cray XMT (3rd generation MTA). Each node
consists of a 500-MHz MTA Threadstorm multithreaded processor that supports
128~streams. The machine uses Cray's proprietary SeaStar2 interconnection
network.

@@

\vspace{0.5ex}
\noindent{\bf IntSort:} This sorting benchmark is taken from the NAS Parallel Benchmark Suite~\cite{Bailey91thenas,nas3.3} and is one on which the Cray XMT's early predecessor once held the world speed record~\cite{TeraRecord}. The largest problem size, class D, ranks two billion uniformly distributed random integers using either a bucket or a counting sort algorithm, depending on the strengths of the system. Bucket sort executes a greater number of loops, but is able to leverage locality and avoid communication completely in the final phase, ranking within buckets. For these reasons, the MPI reference version and our \Grappa implementation use bucket sort. On the other hand, the Cray XMT cannot take advantage of locality, but has an efficient compiler-supported parallel prefix sum, so it performs best using the counting sort algorithm.  The performance metrics for NAS Parallel Benchmarks, including IntSort, are ``millions of operations per second'' (MOPS). For IntSort, this ``operation'' is ranking a single key, so it is roughly comparable to ``GUPS'' or ``TEPS.''

### 2013-07-24 doc/asplos14/overview.tex
fixed todo
Cora->Manta
more small fixes

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} \Grappa system overview\TODO{Cora->Manta}}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

@@

\item [Tasking system.] The tasking system supports lightweight multithreading
to tolerate communication latency and global distributed work-stealing (i.e.,
tasks can be stolen from any node in the system), which provides automated
load balancing. The scheduler oversubscribes to have more worker threads than
required for latency tolerance. By keeping at least four workers per 
core ready to run at all times, the scheduler can prefetch a worker's state
into cache to reduce the chance of stalling on memory accesses during a
context switch.

\item[Communication layer.] The main goal of our communication layer is to
aggregate small messages into large ones. This process is invisible to the
application programmer. Its interface is based on active
messages~\cite{vonEicken92}. Since aggregation and deaggregation of messages
needs to be very efficient, we perform the process in parallel and carefully
use lock-free synchronization operations.

@@

\item[Distributed shared memory.] The DSM system provides fine-grain access to
data anywhere in the system, with delegated operation at the core of its design.
Every piece of global memory is owned by a particular core in the system, and all others may only access that memory by delegating their requests to the owning core.
It supports normal access operations such as
\emph{read\/} and \emph{write\/} as well as synchronizing operations such as
\emph{fetch-and-add\/}~\cite{fetchandadd}. Due to delegation, the memory model offered is similar to what underpins
C/C++~\cite{N2480,N2800}, so it is familiar to programmers. The DSM system
design relies on the lightweight tasking system and communication layer in
order to offer high aggregate random access bandwidth for accessing remote
data.

>>>>

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{figs/system-overview}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} \Grappa system overview}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

@@

\item [Tasking system] The tasking system supports lightweight multithreading
to tolerate communication latency and global distributed work-stealing (i.e.,
tasks can be stolen from any node in the system), which provides automated
load balancing. The scheduler oversubscribes to have more worker threads than
required for latency tolerance. By keeping at least four workers per 
core ready to run at all times, the scheduler can prefetch a worker's state
into cache to reduce the chance of stalling on memory accesses during a
context switch.

\item[Communication layer] The main goal of our communication layer is to
aggregate small messages into large ones. This process is invisible to the
application programmer. Its interface is based on active
messages~\cite{vonEicken92}. Since aggregation and deaggregation of messages
needs to be very efficient, we perform the process in parallel and carefully
use lock-free synchronization operations.

@@

\item[Distributed shared memory] The DSM system provides fine-grain access to
data anywhere in the system, with delegated operation at the core of its design.
Every piece of global memory is owned by a particular core in the system, and all others may only access that memory by delegating their requests to the owning core.
It supports normal access operations such as
\emph{read\/} and \emph{write\/} as well as synchronizing operations such as
\emph{fetch-and-add\/}~\cite{fetchandadd}. Due to delegation, the memory model offered is similar to what underpins
C/C++~\cite{N2480,N2800}, so it is familiar to programmers. The DSM system
design relies on the lightweight tasking system and communication layer in
order to offer high aggregate random access bandwidth for accessing remote
data.

### 2013-07-24 doc/asplos14/tasks.tex
more small fixes

\begin{itemize}

\item {\it deadline task queue}, a priority queue of tasks that are executed according to task-specific deadline constraints;

\item {\it private task queue}, a queue of tasks that must run on
this core and is therefore not subject to stealing;

\item {\it public task queue}, a queue of tasks that are
  waiting to be matched with workers. It is a local partition of a shared
  task pool.

\end{itemize}
Whenever a task yields, the scheduler makes a decision about what
to do next. First, any task in the deadline task queue whose deadline
is imminent is chosen for execution. This queue manages high priority
system tasks, such as periodically servicing communication requests. Second,
the scheduler determines if any workers with running tasks are ready to
execute; if so, one is scheduled. Finally, if no workers are ready to
run, but tasks are waiting to be matched with workers, an idle worker is
woken (or a new worker is spawned), matched with a task, and scheduled.

>>>>

\begin{description}

\item[deadline task queue] a priority queue of tasks that are executed according to task-specific deadline constraints;

\item[private task queue] a queue of tasks that must run on
this core and is therefore not subject to stealing;

\item[public task queue] a queue of tasks that are
  waiting to be matched with workers. It is a local partition of a shared
  task pool.

\end{description}
Whenever a task yields, the scheduler makes a decision about what
to do next. First, any task in the deadline task queue whose deadline
is imminent is chosen for execution. This queue manages high priority
system tasks, such as periodically servicing communication requests. Second,
the scheduler determines if any workers with running tasks are ready to
execute; if so, one is scheduled. Finally, if no workers are ready to
run, but tasks are waiting to be matched with workers, an idle worker is
woken (or a new worker is spawned), matched with a task, and scheduled.
